---
title: 'Exercise 1_2_3: Statistical Software Files'
author: 'Cute Cat'
date: 'Introduction to R for Data Analysis'
---

```{r child = "../config/exercises_setup.Rmd"}
```

Now we are in the second exercise session. Non-flat files were promised, so please load the *GESIS Panel* COVID-19 survey data.

This first part of the exercises only deals with importing data. Later, in the second part, we will turn to labelling and exporting.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Load the *GESIS Panel* COVID-19 survey data.
```

```{block, opts.label = "clues"}
You can either load the *SPSS* or *Stata* file. In both cases, you need the `haven` package.
```

```{r load_gp, solution = TRUE}
library(haven)

gp_covid <-
  read_spss("./data/ZA5667_v1-1-0.sav")
```

In contrast to the flat files, such as CSV, the variables now have labels.

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Print the labels of the first ten variables in the dataset.
```

```{block, opts.label = "clues"}
You can print labels with the function `sjlabelled::get_label(your_data)`, but you have to make sure only to print the first ten variables.
```

```{r print_labels, solution = TRUE, message = FALSE}
library(sjlabelled)

get_label(gp_covid[1:10])
```

Unfortunately, it's all in German. Imagine you are an education researcher, and you are interested in the variable `education_cat`. So you may want to consider translating the variable into English.

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Change the variable label of `education_cat` from "Bildung, kategorisiert" to "Education, categorized".
```

```{block, opts.label = "clues"}
There a two ways: You can either use the function `sjlabelled::set_label()` or do it in a pipe with `sjlabelled::var_labels()`.
```

```{r changevarlabel, solution = TRUE, message = FALSE}
# either
gp_covid$education_cat <- 
  set_label(
    gp_covid$education_cat, 
    label = "Education, categorized"
  )

# or
library(dplyr)

gp_covid <- 
  gp_covid %>% 
  var_labels(
    education_cat = "Education, categorized"
  )

# proof
get_label(gp_covid$education_cat)
```

Your colleague asks you to provide your new data after changing labels and stuff. Unfortunately, she does not use `R` or `SPSS` and asks you to export your data as a *Stata* file.

```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Export your data as a *Stata* file.
```

```{block, opts.label = "clues"}
As in the case of importing you, again, need the `haven` package.
```

```{r exportstata, solution = TRUE, eval = FALSE}
write_stata(gp_covid, "gesis_panel_corona_fancy_panels_final_final.dta")
```