---
title: 'Exercise 3_2_1: A Simple Plot'
author: 'Johannes Breuer & Stefan Jünger'
date: 'Introduction to R for Data Analysis'
---

```{r child = "../config/exercises_setup.Rmd"}
```

Plotting data is fun. We also do it for data exploration (see the section on *Exploratory Data Analysis*). In these exercises, however, we just want to familiarize ourselves with some of the graphic facilities in R.

So let's start with a simple task: creating a simple scatter plot. For this purpose, we first load the *GESIS Panel* COVID-19 survey data.

```{r loadgpcovid, message=FALSE, eval=TRUE}
library(dplyr)
library(haven)

gp_covid <- 
  read_sav(
    "./data/ZA5667_v1-1-0.sav"
  ) %>% 
  sjlabelled::set_na(na = c(-1:-99, 97:98))
```

Now let's plot an actual variable from the dataset.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Create a scatter plot for the variables `political_orientation` and `hzcy001a`.
```

```{block, opts.label = "clues"}
You simply need the `plot()` command.
```

```{r scatter, solution = TRUE}
plot(gp_covid$political_orientation, gp_covid$hzcy001a)
```

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Add some jitter to the scatter plot.
```

```{block, opts.label = "clues"}
You need the `jitter()` function. Have a look at `?jitter` for more details on using the function.
```

```{r jitter, solution = TRUE}
plot(
  jitter(gp_covid$political_orientation, 2), 
  jitter(gp_covid$hzcy001a, 2)
)
```

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Add some color of your choice to the plot.
```

```{block, opts.label = "clues"}
Did you check out the `ColourPicker` add-in for more modern colors?
```

```{r colorscatter, solution = TRUE}
plot(
  jitter(gp_covid$political_orientation, 2), 
  jitter(gp_covid$hzcy001a, 2),
  col = c("#1C86EE", "#FFFFFF", "#FFFFFF")
)
```