---
title: 'Exercise 1_2_3: Statistical Software Files'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
---

```{r child = "../config/exercises_setup.Rmd"}
```

In this set of exercises, we will work with files from statistical software. The first tasks are about importing data, while the later ones are about labelling and exporting.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Import the `.sav` version of the data from the *GESIS Panel Special Survey on the Coronavirus SARS-CoV-2 Outbreak in Germany*.
```

```{block, opts.label = "clues"}
You need the `haven` package for this. The file should be stored in the `data` folder.
```

```{r load_gp, solution = TRUE}
library(haven)

gp_covid <-
  read_spss("./data/ZA5667_v1-1-0.sav")
```

Unlike in flat files, such as CSV, the variables now have labels.

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Print the labels of the first ten variables in the data set.
```

```{block, opts.label = "clues"}
You can use a function from the `sjlabelled` package for this. Remember that you can use [ ] ro subset columns/variables (we only want to print the labels for the first ten variables).
```

```{r print_labels, solution = TRUE, message = FALSE}
library(sjlabelled)

get_label(gp_covid[1:10])
```

Unfortunately, it's all in German. Imagine you are an education researcher working on a publication in English, and you are interested in the variable `education_cat`. So you may want to consider translating the variable into English.

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Change the variable label of `education_cat` from "Bildung, kategorisiert" to "Education, categorized".
```

```{block, opts.label = "clues"}
You can, again, use a frunction from `sjlabelled` for this.
```

```{r changevarlabel, solution = TRUE, message = FALSE}
gp_covid$education_cat <- 
  set_label(
    gp_covid$education_cat, 
    label = "Education, categorized"
  )

get_label(gp_covid$education_cat)
```

Your collaborators ask you to share the data after changing labels and stuff. Unfortunately, they do not use `R` or `SPSS` and, hence, asks you to export your data as a *Stata* file.

```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Export your data as a *Stata* file.
```

```{block, opts.label = "clues"}
The `haven` package provides a function for writing such files that is called and works in a similar way as the corresponding function for importing data in this particular format.
```

```{r exportstata, solution = TRUE, eval = FALSE}
write_stata(gp_covid, "gesis_panel_corona_fancy_panels_final_final.dta")
```