---
title: 'Exercise 2 Data Visualization Part 1 - Solutions'
# title: 'Exercise 2 Data Visualization Part 1'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

After we have tinkered with general plotting in `base R`, and before we learn about alternative plotting methods using `ggplot2`, we want to use this exercise to programmatically play with some `base R` plots. In the previous exercise, we had some repeated tasks to perform. Maybe we can find a more elegant way of creating multiple plots.

Make sure our *GESIS Panel* COVID-19 survey data are loaded:

```{r loadgpcovid, message=FALSE, eval=TRUE}
library(dplyr)
library(haven)

gp_covid <- 
  read_sav(
    "../data/ZA5667_v1-1-0.sav"
  ) %>% 
  sjlabelled::set_na(na = c(-1:-99, 97))
```

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Write a function that automatically creates a table of an input variable and feeds it into a barplot. Moreover, provide an interface to enter the variable name as a character string and the dataset separately by including two commands in the function call.
```

```{block, opts.label = "clues"}
After the function call, use two function arguments with `function (variable, data)` and call the variable within the function with `data[[variable]]`.
```

```{r barplotfunction, solution = TRUE}
barplotter <- function (variable, data) {
  barplot(table(data[[variable]]))
}

barplotter("hzcy001a", gp_covid)
```

Alas, that was not about plotting but rather about programming (sorry!). But now you can use this function and include it in a loop to actually plot stuff!

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Take all four variables from the previous exercise session and feed them in a loop that calls the plotting function. You may want again call `par()` before the loop to have all variables nicely plotted in one graph.
```

```{block, opts.label = "clues"}
Define a character vector with the variable names before you build your graph. For the loop, I'd suggest using a `for()` loop, which iterates over this character vector.
```

```{r barplotfunctionloop, solution = TRUE}
variable_names <- c("hzcy001a", "hzcy002a", "hzcy003a", "hzcy004a")

par(mfrow = c(2, 2))
for (variable in variable_names) {
  barplotter(variable, gp_covid)
}
```

Before we later start with the other exercises, you may want to consider to clean your graphics device with `dev.off()`

```{r cleandevice , message=FALSE}
dev.off()
```