---
title: 'Exercise 2 Confirmatory Analysis - Solutions'
# title: 'Exercise 2 Confirmatory Analysis'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

In the last exercise, we made use of some regression models for our analysis. That was a good start, but publications need tables of results and fancy visualizations (e.g., for model predictions).

Again, we need to load the *GESIS Panel* COVID-19 survey data and also - just to make sure - re-run the regression models:

```{r loadgpcovid, message=FALSE}
library(dplyr)
library(haven)
library(sjlabelled)

gp_covid <-
  read_sav(
    "../data/ZA5667_v1-1-0.sav"
  ) %>% 
  set_na(na = c(-1:-99, 97)) %>% 
  mutate(
    likelihood_hospital = hzcy003a,
    age_cat = as.factor(age_cat),
    likelihood_hospital_cut =
      ifelse(
        likelihood_hospital > median(likelihood_hospital, na.rm = TRUE),
        1,
        0
      )
  ) %>% 
  remove_all_labels()

serious_logistic_regression <- 
  glm(
    likelihood_hospital_cut ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid,
    family = binomial(link = "logit")
  )

serious_cauchit_regression <- 
  glm(
    likelihood_hospital_cut ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid,
    family = binomial(link = "cauchit")
  )
```

In the lecture, we started with the visualization of predictions. However, it is good practice to first build a regression table that shows, for example, the statistical significance of regression coefficients.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Build a `stargazer` regression table of both models to compare them. Use the option `type = "text"` to have it printed to the console. Moreover, choose a table style for a journal of your choice.
```

```{block, opts.label = "clues"}
You can find an overview of all supported styles with ``?stargazer::`list of supported styles` ``.
```

```{r stargazer, solution = TRUE}
library(stargazer)

stargazer(
  serious_logistic_regression,
  serious_cauchit_regression,
  type = "text",
  style = "asr"
)
```

You know what: reviewer 2 is "post-significant" (at least a little bit). He requests that you should please use confidence intervals at a level of `.93` (which is now all the rage in his field).

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Rebuild your regression table and add the confidence intervals with the requested level.
```

```{block, opts.label = "clues"}
Have a look at the help page for all possible options `??stargazer`. Note that it is possible be that CIs are not shown because of the custom style you choose.
```

```{r stargazer2, solution = TRUE}
library(stargazer)

stargazer(
  serious_logistic_regression,
  serious_cauchit_regression,
  type = "text",
  ci = TRUE,
  ci.level = .93
)
```

The previous exercise might also show that integrated packages like `stargazer` have limits. Because of these limits, standardized procedures of gathering results as offered by the `broom` package are quite attractive since packages like, e.g., [`huxtable`](https://hughjonesd.github.io/huxtable/) 'understand' this standard format.

We'll leave regression tables now and turn to plotting the results (as reviewer 2 wants you to do that after the third review round)...

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Draw a prediction plot of the variable `age_cat` for both models using the `sjPlot` package. Also, try to plot them side by side.
```

```{block, opts.label = "clues"}
You can use the `patchwork` package for creating the combined plot.
```

```{r predictionplot, solution = TRUE}
library(sjPlot)
library(patchwork)

logit_plot <- 
  plot_model(
    serious_logistic_regression,
    type = "pred",
    terms = "age_cat"
    )

cauchit_plot <- 
  plot_model(
    serious_cauchit_regression,
    type = "pred",
    terms = "age_cat"
    )

logit_plot | cauchit_plot
```

Reviewer 2 is happy now. But Reviewer 1, who is more trusted by the editor, thinks estimating predictions based on 'real' average marginal effects (AME) is more cutting-edge. She also feels that you have proven that choosing a logit or a cauchit link doesn't make a difference. Therefore she'd prefer to see only one plot for the cauchit model as this is not a methodological paper.

```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Draw a prediction plot of the variable `age_cat` only for the cauchit model, but this time based on AME.
```

```{block, opts.label = "clues"}
The `margins` package and its `cplot()` function are your friends here.
```

```{r predictionplot2, solution = TRUE}
library(margins)

cplot(
  serious_cauchit_regression,
  "age_cat",
  what = "prediction"
)
```

Bonus: You may have seen that `cplot()` outputs the actual prediction data automatically. This was annoying in the slides, but you could use this output to create a plot with `ggplot2`. *Note*: When you set the argument `draw = FALSE` only the data are created.

```{block, box.title = "BONUS", box.body = list(fill = "white"), box.icon = "fa-star"}
Draw a prediction plot of the variable `age_cat` only for the cauchit model based on AME using `ggplot2`.
```

```{block, opts.label = "clues"}
1. The column `xvals` is stored as character. You should convert it to numeric.
2. There's a `geom` called `geom_errorbar` which you can use for the confidence intervals
```

```{r predictionplot3, solution = TRUE}
library(ggplot2)

prediction_data <-
  cplot(
  serious_cauchit_regression,
  "age_cat",
  what = "prediction",
  draw = FALSE
) %>% 
  dplyr::mutate(
    xvals = as.numeric(xvals)
  )

ggplot(
  prediction_data,
  aes(x = xvals, y = yvals)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = .1) +
  theme_bw()
```