---
title: "Introduction to R for Data Analysis"
subtitle: "Data Types, Import & Export"
author: "Johannes Breuer & Stefan Jünger"
date: "2021-08-02"
presenter: Stefan
---
layout: true 

```{r child = "../config/sessions_setup.Rmd"}
```

---

## Getting data into `R`
Thus far, we've already learned what `R` and `RStudio` are. This course is about starting to use `R` and feeling prepared to use it for statistical analyses. There's one essential prerequisite:

.center[**We need data!**]

```{r, out.width = "50%", echo = FALSE}
woRkshoptools::include_picture("./import_data.png")
```

---

## Content of this session 

- What are `R`'s internal data types?
- How to work with different data types?
- How to import data in different formats?
- How to export data in different formats

---

## Data we use in this course

During the course, we use several different data sets. Especially in this session, where we apply different importing functions, we quite a few data sets, from data about the Titanic to data about unicorns. However, we will also use data that are more interesting for social and behavioral scientists.

---

## It boils all down to...

.pull-left[
**How your data are stored (data types)**
- 'Numbers' (Integers & Doubles)
- Character Strings
- Logical
- Factors
- ...
- There's more, e.g., expressions, but let's leave it at that
]

.pull-right[
**Where your data are stored (data formats)**
- Vectors
- Matrices
- Arrays
- Data frames / Tibbles
- Lists
]

.footnote[https://www.stat.berkeley.edu/~nolan/stat133/Fall05/lectures/DataTypes4.pdf]

---

## Numeric data
.small[
*Integers* are values without a decimal value. To be explicit in `R` in using them, you have to place an `L` behind the actual value.

```{r integer-1}
1L
```

By contrast, *doubles* are values with a decimal value.

```{r integer-2}
1.1
```

We can check data types by using the `typeof()` function.

```{r typeof}
typeof(1L)
typeof(1.1)
```
]

---

## Character strings
At first glance, a *character* is a letter somewhere between a-z. *String* in this context might mean that we have a series of characters. However, numbers and other symbols can be part of a *character string*, which can then be, e.g., part of a text. In `R`, character strings are wrapped in quotation marks.

```{r string}
"Hi. I am a character string, the 1st of its kind!"
```

*Note*: There are no values associated with the content of character strings unless we change that, e.g., with factors.

---

## Factors

If you're a *Stata* (or *SPSS*) user, you may already be  familiar with factors. Factors are data types that assume that their values are not continuous, e.g., as in ordinal or nominal data.

```{r factor}
factor(1.1)
factor("Hi. I am a character string, the 1st of its kind!")
```

Factors take numeric data or character strings as input as they simply convert them into so-called levels. This concept may be a little bit abstract for the time being. It's just essential to have heard about them before you learn more about them.

---

## Logical values

Logical values are basically either `TRUE` or `FALSE` values. These values are produced by making logical requests on your data.

```{r logical}
2 > 1
2 < 1
```

Logical values are at the heart of creating loops. For this purpose, however, we need more logical operators to request `TRUE` or `FALSE` values.

---

## Logical operators

There are quite a few logical operators in `R`:

.pull-left[
- `<` 	less than
- `<=` 	less than or equal to
- `>` 	greater than
- `>=` 	greater than or equal to
- `== `	exactly equal to
- `!=` 	not equal to
]

.pull-right[
- `!x` 	Not x
- `x | y` 	x OR y
- `x & y `	x AND y
- `isTRUE(x)` 	test if X is TRUE 
- `isFALSE(x)` 	test if X is FALSE 
]

.footnote[https://www.statmethods.net/management/operators.html]

Moreover, there are some more `is.PROPERTY_ASKED_FOR()` functions, such as `is.numeric()`, which also return `TRUE` or `FALSE` values.

---

## `R`'s data formats 

`R`'s different data types can be put into 'containers'.

```{r containers, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("9213.1526125966.png")
```

.footnote[https://devopedia.org/r-data-structures]

---

## Vectors

Vectors are built by enclosing your content with `c()` ("c" for "concatenate")

```{r vectors}
numeric_vector   <- c(1, 2, 3, 4)
character_vector <- c("a", "b", "c", "d")

numeric_vector
character_vector
```

Vectors are really like vectors in mathematics. Initially, it doesn't matter if you look at them as column or row vectors.

---

## ...but it matters when you combine vectors

Using the function `cbind()` or `rbind()` you can either combine vectors column-wise or row-wise. Thus, they become matrices.

```{r combine-vectors}
cbind(numeric_vector, character_vector)
rbind(numeric_vector, character_vector)
```

.small[
*Note*: The numeric values are coerced into strings here.
]

---

## Matrices

Matrices are the basic rectangular data format in R.

```{r matrix}
fancy_matrix <- matrix(1:16, nrow = 4)

fancy_matrix
```

You cannot store multiple data types, such as strings and numeric values in the same matrix.  Otherwise, your data will get coerced to a common type, as seen in the previous slide. This is something that happens already within vectors:

```{r vector-coercion}
c(1, 2, "evil string")
```

---

## Data frames

While matrices are used, e.g.,--\*drumroll\*-- for matrix operations, data frames resemble more the data formats most of you are probably already familiar with. We can build data frames by hand as here:

.tinyish[
```{r data-frames, eval = FALSE}
library(randomNames) # a name generator package

fancy_data <-
  data.frame( 
    who = 
      randomNames(n = 10, which.names = "first"),
    age = 
      sample(14:49, 10, replace = TRUE), # you see what we are doing here?   
    salary_2018 = 
      sample(15:100, 10, replace = TRUE),  
    salary_2019 = 
      sample(15:100, 10, replace = TRUE)
  )
 
fancy_data
```
]

.right[`r emo::ji("left_arrow_curving_right")`]

---
class: middle
```{r ref.label = "data-frames", echo = FALSE}
```

---

## Tibbles

.pull-left[
Tibbles are basically just `R data.frames` but nicer.

- only the first ten observations are printed
  - the output is tidier!
- you get some additional metadata about rows and columns that you would normally only get when using `dim()` and other functions

You can check the [tibble vignette](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html) for technical details.
]

.pull-right[
```{r tibble-pic, echo = FALSE, out.width="60%"}
woRkshoptools::include_picture("./tibble.png")
```
]

---

## Tibble conversion



```{r as_tibbles}
library(tibble)
as_tibble(fancy_data)
```

---

## One last type you should know: lists

Lists are perfect for storing numerous and potentially diverse pieces of information in one place.

```{r fancy_list, eval = FALSE}
fancy_list <- 
  list(
    numeric_vector,
    character_vector,
    fancy_matrix,
    fancy_data
  )

fancy_list
```

.right[`r emo::ji("left_arrow_curving_right")`]

---
class: middle
.tinyish[
```{r ref.label = "fancy_list", echo = FALSE}
```
]

---

## Nested lists

```{r fancy_nested_list, eval = FALSE}
fancy_nested_list <-
  list(
    fancy_vectors = list(numeric_vector, character_vector),
    data_stuff = list(fancy_matrix, fancy_data)
  )

fancy_nested_list
```

.right[`r emo::ji("left_arrow_curving_right")`]

---
class: middle
.tinyish[
```{r ref.label = "fancy_nested_list", echo = FALSE}
```
]

---

## Accessing elements by index

Generally, the logic of `[index_number]` is used in `R` to access only a subset of information in an object, no matter if we have vectors or data frames.

Say, we want to extract the second element of our `character_vector` object, we could do that like this:

```{r}
character_vector[2]
```

---

## More complicated cases: matrices

Matrices can have more dimensions, often you want information from a specific row and column.

```{r eval = FALSE}
a_wonderful_matrix[number_of_row, number_of_column]
```

*Note*: You can do the same indexing with `data.frame`s. We will talk more about this in the session on *Data Wrangling Basics*.

---

## Matrices and subscripts (as in mathematical notation)

Identifying rows, columns, or elements using subscripts is similar to matrix  notation:

```{r eval = FALSE}
fancy_matrix[, 4] # 4th column of matrix
fancy_matrix[3,] # 3rd row of matrix
fancy_matrix[2:4, 1:3] # rows 2,3,4 of columns 1,2,3 
```

It's really like in math, and you can perform standard mathematical operations, such as matrix multiplications.

```{r}
fancy_matrix[2:4, 1:3] %*% fancy_matrix[1:3, 2:4]
```

---

## The case of data frames

A nice feature of `data.frames` or `tibbles` is that their columns are names, just as variable names in ordinary data. It would be cumbersome to use index numbers to extract a specific column/variable, right? Do not fear:

```{r}
fancy_data$who
```

Just place a `$`-sign between the data object and the variable name.

---

## `[]` in data frames

Sometimes we also have to rely on character strings as input information, e.g., for iterating over data. We can also use `[]` to access variables by name. 

.pull-left[ 
Not only this way:

```{r}
fancy_data[1]
```
]

.pull-right[
But also this way:

```{r}
fancy_data["who"]
```
]
 
---

## Difference between `[]` and `[[]]`

```{r hadley-tweet, echo=FALSE, out.width="80%"}
tweet_screenshot("https://twitter.com/hadleywickham/status/643381054758363136",
            theme = "dark")
```
https://twitter.com/hadleywickham/status/643381054758363136

---

## Data frame check 1, 2, 1, 2!

Once you start working with data in `R` a good first thing to do is to have a quick look at them. The most high-level information you can get is about the object type and its dimensions.

.small[
```{r class-dim}
# object type
class(fancy_data)

# number of rows and columns
dim(fancy_data)

# number of rows
nrow(fancy_data)

# number of columns
ncol(fancy_data)
```
]

---

## Data frame check 1, 2, 1, 2!

You can also print the first 6 lines of the data frame with `head()`. You can easily change the number of lines by providing the number as the second argument to the `head()` function.

```{r head}
head(fancy_data, 3)
```

---

## Data frame check 1, 2, 1, 2!

If we want some more (detailed) information about the data set or object, we can use the `base R` function `str()`.

```{r str}
str(fancy_data)
```

---

## Data frame check 1, 2, 1, 2!

If you want to have a look at your full data set, you can use the `View()` function. In *RStudio*, this will open a new tab in the source pane through which you can explore the data set (including a search function). You can also click on the small spreadsheet symbol on the right side of the object in the environment tab to open this view. 

```{r view, eval = F}
View(fancy_data)
```

```{r, view-pic, out.width = "65%", echo = F}
woRkshoptools::include_picture("./rstudio_view.png")
```

---

## Viewing and changing names

We can print all names of an object using the `names()` function...

```{r receiving-names}
names(fancy_data)
```

...and we can also change names with it.

```{r changing-names}
names(fancy_data) <- c("name", "salary_2018", "salary_2019")

names(fancy_data)
```

However, there are more flexible ways of doing this as we will see in the session on *Data Wrangling Basics* tomorrow.

---

class: center, middle


# [Exercise](https://jobreu.github.io/r-intro-gesis-2021/exercises/Exercise_1_2_1_Data_Types.html) time `r emo::ji("weight_lifting_woman")``r emo::ji("muscle")``r emo::ji("running_man")``r emo::ji("biking_man")`

## [Solutions](https://jobreu.github.io/r-intro-gesis-2021/solutions/Exercise_1_2_1_Data_Types.html)


---

## GESIS Panel Data on the Coronavirus Outbreak
.left-column[
```{r echo = FALSE}
woRkshoptools::include_picture("./gesis_panel_logo_web.jpg")
```
]

.right-column[
For most of the examples and exercises in this course we will use the [Public Use File (PUF) of the GESIS Panel Special Survey on the Coronavirus SARS-CoV-2 Outbreak in Germany](https://www.gesis.org/gesis-panel/coronavirus-outbreak/public-use-file-puf). You can [download the data set in different formats as well as the codebook and the questionnaire (in German) from the *GESIS* Data Archive](https://search.gesis.org/research_data/ZA5667) (note: you need to have/create a user account).

The *GESIS Panel* website provides [detailed documentation](https://www.gesis.org/gesis-panel/documentation), including a [cheatsheet](https://www.gesis.org/fileadmin/upload/GESIS_Panel/Cheatsheet/gesis_panel_cheatsheet.pdf).
]

---

## Gapminder Data
.left-column[ 
```{r echo = FALSE}
woRkshoptools::include_picture("./gapminder_logo.png")
```
]

.right-column[
We will also use [data from *Gapminder*](https://www.gapminder.org/data/). During the course and the exercises, we work with data we have downloaded from their website. There also is an `R` package that bundles some of the *Gapminder* data: `install.packages("gapminder")`.

This `R` package provides ["[a]n excerpt of the data available at Gapminder.org. For each of 142 countries, the package provides values for life expectancy, GDP per capita, and population, every five years, from 1952 to 2007."](https://cran.r-project.org/web/packages/gapminder/index.html)
]

---

## How to use the data in general

To code along and be able to do the exercises, you should store the data files for the *GESIS Panel Special Survey on the Coronavirus SARS-CoV-2 Outbreak in Germany* in a folder called `data` in the same folder as the other materials for this course.

---

## `R` is data-agnostic
```{r, echo = FALSE, out.width = "65%"}
woRkshoptools::include_picture("./Datenimport.PNG")
```

---

## Data formats & packages
.pull-left[
**What you will learn**
- Getting the most common data formats into `R`
  - e.g., CSV, *Stata*, *SPSS*, or *Excel* spreadsheets
- Using the different methods of doing that
- We will rely a lot on packages and functions from the `tidyverse` instead of using `base R`
]

.pull-right[
**What you won't learn**
- Getting old & obscure binary data formats into `R`
  - ... although [that is possible](https://cran.r-project.org/doc/manuals/r-release/R-data.html)
]

---

## Before writing any code: *RStudio* functionality for importing data
You can use the *RStudio* GUI for importing data via `Environment - Import data set - Choose file type`.

```{r, echo = FALSE}
woRkshoptools::include_picture("./rstudio_import.PNG")
```

---

## Where to find data

**Browse Button in `RStudio`**
```{r, echo = FALSE, out.width="75%"}
woRkshoptools::include_picture("./importBrowse.PNG")
```

**Code preview in `Rstudio`**
```{r, echo = FALSE, out.width="75%"}
woRkshoptools::include_picture("./codepreview.PNG")
```

---

## Honestly, after some time you will write the code directly

.center[
```{r, echo = FALSE}
woRkshoptools::include_picture("./coding_cat.gif")
```
.footnote[[Source](https://media.giphy.com/media/LmNwrBhejkK9EFP504/source.gif)]
]

---

## Honestly, after some time you will write the code directly

.center[
```{r, echo = FALSE}
woRkshoptools::include_picture("./hadley-typing.gif")
```
[Source](https://tenor.com/view/hadley-wickham-rstats-typing-rcode-gif-11365139)
]

---

## Simple vs. not so simple file formats

Basic file formats, such as CSV (comma-separated value file), can directly be imported into `R`
- they are 'flat'
- few metadata
- basically text files

Other file formats, particularly the proprietary ones, require the use of additional packages
- they are complex
- a lot of metadata (think of all the labels in an *SPSS* file)
- they are binary (1110101)


---

## File formats wars

```{r echo = FALSE, out.width = "30%"}
woRkshoptools::include_picture("./norm_normal_file_format.png")
```
https://xkcd.com/2116/

---

## Disclaimer

**In the following slides, we'll jump right into importing data. We use a lot of different packages for this purpose, and you don't have to remember everything. It's just for making a point of how agnostic `R` actually is regarding the file type. Later on, we will dive more into the specifics of importing.**

---

## Importing a CSV file using `base R`

```{r loadtitanic, eval = FALSE}
titanic <- read.csv("./data/titanic.csv")

titanic
```

.tinyish[
```{r loadtitanic-print, echo = FALSE}
titanic <- read.csv("./data/titanic.csv")

titanic
```
]

---

## A `readr` example: `CSV` files
```{r readr-example}
library(readr)

titanic <- read_csv("./data/titanic.csv")
```

---
class: middle

.tinyish[
```{r readr-example-output, echo = TRUE}
titanic
```
]

Note the column specifications: `readr` 'guesses' them based on the first 1000 observations (we will come back to this later).

---

## Importing *Excel* files with `readxl`

```{r readxl-example}
library(readxl)

unicorns <- read_xlsx("./data/observations.xlsx")
```

No output `r emo::ji("frowning_face")`

---
class: middle
```{r readxl_example_output, echo = TRUE}
unicorns
```

---

## *Stata* files with `haven`

```{r read-stata-example-display, eval = FALSE}
library(haven)

gp_covid <- 
  read_stata("./data/ZA5667_v1-1-0_Stata14.dta")

gp_covid
```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle


```{r ref.label = "read-stata-example-display", echo = FALSE}
```


---

## *SPSS* files with `haven` 

The `haven` package also offers the function `read_spss()` for importing *SPSS* files.

The package also offers capabilities for handling *SPSS*-defined missing values by setting the option `user_na = TRUE` (default is `FALSE`).

*Note*: The [`sjlabelled` package](https://cran.r-project.org/web/packages/sjlabelled/index.html) can also be used for [working with user-defined missings from *SPSS* files](https://cran.r-project.org/web/packages/sjlabelled/vignettes/intro_sjlabelled.html).

**We will come back to *Stata* and *SPSS* files in a bit as they represent a specific file format in `R`: labelled data.**

---

## Other data import options

These were just some very first examples of applying functions for data import from the different packages. There are many more...

.pull-left[
`readr`
- `read_csv()`
- `read_tsv()`
- `read_delim()`
- `read_fwf()`
- `read_table()`
- `read_log()`
]

.pull-right[
`haven`
- `read_sas()`
- `read_spss()`
- `read_stata()`
]

Not to mention all the helper functions and options. For example, we can define the cells to read from an *Excel* file by specifying the option `range = "C1:E4"` in `read_excel()`

---

## Data type specifications for `tibbles` 

- characters
  - indicated by `<chr>`
  - specified by `col_character()`
- integers
  - indicated by `<int>`
  - specified by `col_integer()`
- doubles
  - indicated by `<dbl>`
  - specified by `col_double()`
- factors
  - indicated by `<fct>`
  - specified by `col_factor()`
- logical
  - indicated by `<lgl>`
  - specified by `col_logical()`

---

## Changing variable types

As mentioned before, `read_csv` 'guesses' the variable types by scanning the first 1000 observations. **NB**: This can go wrong!

Luckily, we can change the variable type...
- before/while loading the data
- and after loading the data

---

## While loading the data in `read_csv`

```{r readr-example-col-change, eval = FALSE}
titanic <-
  read_csv(
    "./data/titanic.csv",
    col_types = cols(
      PassengerId = col_double(),
      Survived = col_double(),
      Pclass = col_double(),
      Name = col_character(),
      Sex = col_character(),
      Age = col_double(),
      SibSp = col_double(),
      Parch = col_double(),
      Ticket = col_character(),
      Fare = col_double(),
      Cabin = col_character(),
      Embarked = col_character()
    )
  )

titanic
```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle

```{r ref.label = "readr-example-col-change", echo = FALSE}
```

---

## While loading the data in `read_csv`

```{r readr-example-col-changeD, eval = FALSE}
titanic <-
  read_csv(
    "./data/titanic.csv",
    col_types = cols(
      PassengerId = col_double(),
      Survived = col_double(),
      Pclass = col_double(),
      Name = col_character(),
      Sex = col_factor(), # This one changed!
      Age = col_double(),
      SibSp = col_double(),
      Parch = col_double(),
      Ticket = col_character(),
      Fare = col_double(),
      Cabin = col_character(),
      Embarked = col_character()
    )
  )

titanic
```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle

```{r ref.label = "readr-example-col-changeD", echo = FALSE}
```

---

## After loading the data

```{r readr_example_col_changeD_after, eval = FALSE, echo = TRUE}
titanic <-
  type_convert(
    titanic,
    col_types = cols(
      PassengerId = col_double(),
      Survived = col_double(),
      Pclass = col_double(),
      Name = col_character(),
      Sex = col_factor(),
      Age = col_double(),
      SibSp = col_double(),
      Parch = col_double(),
      Ticket = col_character(),
      Fare = col_double(),
      Cabin = col_character(),
      Embarked = col_character()
    )
  )
```

---

## Beyond flat files: labelled data

A lot of data comes in some sort of flat file format, such as `CSV`. In the social sciences, however, we often deal with proprietary file formats, such as *SPSS*'s `.sav` or *Stata*'s `.dta` files. 

What these data typically include are labels. These labels are used to describe variables or variable values. They comprise some specific metadata inherent in these proprietary file formats. 

*If you were able to travel back ten years in time and ask an `R` geek, she'd say that you cannot use labels in R. You'd either have to import, e.g., value labels as character strings or use their codes as factors. However, these days...*

---

## Not being able to use labelled data is a thing of the past

Nowadays, if you use the `haven` package, labels are built-in for the corresponding file types. For example:

```{r}
gp_covid <-
  haven::read_sav("./data/ZA5667_v1-1-0.sav")

gp_covid["age_cat"]
```

---

## Advantages of using labelled data

One could rejoice in not having to use a codebook anymore, just like in *SPSS* (although just looking at code output for glimpsing feels much more... data-geeky). 

An advantage is definitely that you can potentially re-use the labels in figures and plots, and some `R` packages do that automatically, such as the [`sjPlot`](https://strengejacke.github.io/sjPlot/) package.

In addition, when you exchange your data with colleagues who do not use `R` or when you plan to publish your data (which you always should if that is possible), being able to export data you have manipulated in `R` in different formats is great.

**However, be aware of the missing values hell that you may enter due to different missing value definitions in *Stata* and *SPSS*.**

---

## Getting labels

For variables:

```{r}
sjlabelled::get_label(gp_covid$age_cat)
```

For values:

.tinyish[
```{r}
sjlabelled::get_labels(gp_covid$age_cat)
```
]

---

## Setting labels: Variables

```{r}
gp_covid$age_cat <- 
  sjlabelled::set_label(gp_covid$age_cat, label = "Age, categorized")

sjlabelled::get_label(gp_covid$age_cat)
```

---

## Setting labels: Values
.tinyish[
```{r}
gp_covid$age_cat <- 
  sjlabelled::set_labels(
    gp_covid$age_cat,
    labels = 
      c(
        "<=25 years", "26 to 30 years", "31 to 35 years", "36 to 40 years",
        "41 to 45 years", "46 to 50 years", "51 to 60 years", 
        "61 to 65 years", "66 to 70 years", ">=71 years"
      )
  )

sjlabelled::get_labels(gp_covid$age_cat)
```
]

---

class: center, middle

# [Exercise](https://jobreu.github.io/r-intro-gesis-2021/exercises/Exercise_1_2_2_Flat_Files.html) time `r emo::ji("weight_lifting_woman")``r emo::ji("muscle")``r emo::ji("running_man")``r emo::ji("biking_man")`

## [Solutions](https://jobreu.github.io/r-intro-gesis-2021/solutions/Exercise_1_2_2_Flat_Files.html)

---

## Exporting data

Sometimes our data have to leave `R`, for example, if we....
- share data with colleagues who do not use `R`
- want to continue where we left off
  - particularly if data wrangling took a long time
  
For such purposes, we also need a way to export our data.

All of the packages we have discussed in this session also have designated functions for that.

```{r, out.width = "50%", echo = FALSE}
woRkshoptools::include_picture("./export_data.png")
```

---

## Examples: CSV and Stata files
```{r export_csv, eval=FALSE}
write_csv(titanic, "titanic_own.csv")
```

```{r export_stata, eval=FALSE}
write_dta(titanic, "titanic_own.dta")
```

---

## `R`'s native file formats

If you plan to continue to work with `R` (something we would always recommend `r emo::ji("stuck_out_tongue_winking_eye")`), there are at least two native 'file formats' to choose from. The advantage of using them is that they are compressed files, so that they don't occupy unnecessarily large disk space. These two formats are `.Rdata`/`.rda` and `.rds`.
The key difference between them is that `.rds` can only hold one object, whereas `.Rdata`/`.rda` can also be used for storing several objects in one file. A helpful property of 

---

## `.Rdata`/`.rda`

Saving

```{r, eval = FALSE}
save(mydata, file = "mydata.RData")
```

Loading
```{r eval=FALSE}
load("mydata.RData")
```

---

## `.rds`

Saving
```{r, eval = FALSE}
saveRDS(mydata, "mydata.rds")
```

Loading
```{r eval=FALSE}
mydata <- readRDS("mydata.rds")
```
  
*Note*: A nice property of `saveRDS()` is that just saves a representation of the object, which means you can name it whatever you want when loading.

---

## Saving just everything

If you have not changed the General Global Options in *RStudio* as suggested in the *Getting Started* session, you may have noticed that, when closing *Rstudio*, by default, the programs asks you whether you want to save the workspace image. 

```{r, out.width = "50%", echo = FALSE}
woRkshoptools::include_picture("./save_image.png")
```

You can also do that whenever you want using the `save.image()` function:

```{r, eval = FALSE}
save.image(file = "my_fancy_workspace.RData")
```

.small[
*Note*: As we've said before, though, this is not something we'd recommend as a worfklow. Instead, you should (explicitly and separately) save your `R` scripts and data sets (in appropriate formats).
]

---

## Additional packages

Besides `readr`, `haven` and `readxl`, there also are some other packages that facilitate importing specific data types as tibbles:

- [`sjlabelled`](https://cran.r-project.org/web/packages/sjlabelled/index.html) for labelled data, e.g., from *SPSS* or *Stata*

- [`sf`](https://github.com/r-spatial/sf) for geospatial data

---

## Other packages for data import

For data import (and export) in general, there are even more options, such as... 

- `base` R

- the [`foreign` package](https://cran.r-project.org/web/packages/foreign/index.html) for *SPSS* and *Stata* files

- [`data.table`](https://cran.r-project.org/web/packages/data.table/index.html) or [`fst`](https://www.fstpackage.org/) for large data sets

- [`jsonlite`](https://cran.r-project.org/web/packages/jsonlite/index.html) for `.json` files

- [`datapasta`](https://github.com/MilesMcBain/datapasta) for copying and pasting data into tribbles (e.g., from websites, *Excel* or *Word* files)

---

## Reminder regarding file paths

In general, you should avoid using absolute file paths to maintain your code reproducibly and future-proof. We already talked about this in the introduction, but this is particularly important for importing and exporting data.

As a reminder: Absolute file paths look like this (on different OS):

```{r eval = FALSE}
# Windows
load("C:/Users/cool_user/data/fancy_data.Rdata")

# Mac
load("/Users/cool_user/data/fancy_data.Rdata")

# GNU/Linux
load("/home/cool_user/data/fancy_data.Rdata")
```

---

## Use relative paths

Instead of using absolute paths, it is recommended to use relative file paths. The general principle here is to start from a directory where your current script currently exists and navigate to your target location. Say we are in the "C:/Users/cool_user/" location on a Windows machine. To load your data, we would use:

```{r eval = FALSE}
load("./data/fancy_data.Rdata")
```

If we were in a different folder, e.g., "C:/Users/cool_user/cat_pics/mittens/", we would use:

```{r eval = FALSE}
load("../../data/fancy_data.Rdata")
```

---

class: center, middle

Please first download the [Public Use File (PUF) of the GESIS Panel Special Survey on the Coronavirus SARS-CoV-2 Outbreak in Germany](https://search.gesis.org/research_data/ZA5667) as .sav, .dta, and .csv file.

# [Exercise](https://jobreu.github.io/r-intro-gesis-2021/exercises/Exercise_1_2_3_Statistical_Software_Files.html) time `r emo::ji("weight_lifting_woman")``r emo::ji("muscle")``r emo::ji("running_man")``r emo::ji("biking_man")`

## [Solutions](https://jobreu.github.io/r-intro-gesis-2021/solutions/Exercise_1_2_3_Statistical_Software_Files.html)