---
title: "Introduction to R for Data Analysis"
subtitle: "Data Wrangling Advanced"
author: "Johannes Breuer & Stefan Jünger"
date: "2021-08-03"
presenter: Stefan
---
layout: true 

```{r child = "../config/sessions_setup.Rmd"}
```

---

## Data wrangling continued `r ji("cowboy_hat_face")`

While in the last sessions we focused on the bread-and-butter tasks of the data preparation business, in this part we will focus on the more 'programmy' side of things.

- altering the content of a whole set of variables
- conditional variable transformation
- formulating logical requests to our data
- writing loops

---

class: middle

**We will largely remain in the world of the `tidyverse` since it makes the steps of wrangling data so transparent and straightforward. However, we will also show that you can easily combine **

---

## Load the data

Again, we will work with the *Public Use File (PUF) of the GESIS Panel Special Survey on the Coronavirus SARS-CoV-2 Outbreak in Germany* as `.csv` file.

```{r load-gesis-panel-data}
gp_covid <- read_csv2("./data/ZA5667_v1-1-0.csv")
```

---

---

## Quickly define missing values

```{r set-missings}
library(sjlabelled)

gp_covid <-
  gp_covid %>%
  set_na(na = c(-99, -77, -33, 98))
```

---

## Variables of interest

Say, we are interested in the (dis)trust towards several authorities during the Corona crisis. There are 9 items on this topic. Let's create some quick on 3 of them.

```{r quick-table}
table(gp_covid$hzcy044a)
table(gp_covid$hzcy047a)
table(gp_covid$hzcy052a)
```

What if we want to conduct some data reduction method (e.g., PCA) and need the variables in reverse order for interpretation purposes?

---

## Recode data **across** defined variables 

The `dplyr` package provides a (new) handy tool to exactly this: `across()`. This function can be used to apply another function to multiple variables at once.

```{r recode-across}
gp_covid <- 
  gp_covid %>% 
  mutate(
    across(
      hzcy044a:hzcy052a,
      ~recode(
        .x,
        `5` = 1, # `old value` = new value
        `4` = 2,
        `2` = 4,
        `1` = 5
      )
    )
  )
```

---

class: middle

```{r quick-table-2}
table(gp_covid$hzcy044a)
table(gp_covid$hzcy047a)
table(gp_covid$hzcy052a)
```

---

## Using `across()` logical conditions

Sometimes we are interested in variables that meet certain conditions. For example, for an anylsis, we want to z-standardize all numeric variables in a dataset. Let's create a temporary subset of our data to exemplify such efforts.

```{r create-tmp-dataset}
gp_covid_tmp <-
  gp_covid %>% 
  select(doi, hzcy044a:hzcy052a)

gp_covid_tmp %>% 
  sample_n(5)  # randomly sample 5 cases from the df
```

---

## z-standardize all numeric variables

The `base R` function to z-standardize a variable is `scale()`.

```{r scale-vars}
gp_covid_tmp <-
  gp_covid_tmp %>% 
  mutate(
    across(
      is.numeric,
      ~scale(.x)
    )
  )

gp_covid_tmp %>% 
  sample_n(5)
```

---

## `dplyr::across()`

```{r, across-cartoon, out.width = "95%", echo = F}
include_picture("across_blank.png")
```
<small><small>Artwork by [Allison Horst](https://github.com/allisonhorst/stats-illustrations)</small></small>

---

## Aggregate variables across rows

Something we might want to do for our analyses is to create aggregate variables, such as sum or mean scores for a set of items.
As `dplyr` operations are applied to columns, whereas such aggregations relate to rows (i.e., respondents), we need to make use of the function `rowwise()`. Say, for example, we want to compute a sum score for all measures that respondents have reported to engage in to prevent an infection with or the spread of the Corona virus.

```{r rowwise-sum-display, eval = F}
gp_covid <- 
  gp_covid %>% 
  rowwise() %>% #<<
  mutate(
    sum_measures = 
      sum(
        c_across(hzcy044a:hzcy052a),
        na.rm = TRUE
      )
  ) %>% 
  ungroup()
```

---

## Aggregate variables

```{r rowwise-sum}
gp_covid <- 
  gp_covid %>% 
  rowwise() %>% #<<
  mutate(
    sum_measures = 
      sum(
        c_across(hzcy044a:hzcy052a),
        na.rm = TRUE
      )
  ) %>% 
  ungroup()
```

Three things to note here: 

1. `c_across()` is a special version of `across()`for rowwise operations. 

2. We use the `ungroup()` function at the end to ensure that `dplyr` verbs will operate the default way when we further work with the `gpc` object. We do not cover grouping in this course (which is especially valuable for summarizing data), but you can check out the [documentation for `group_by()`](https://dplyr.tidyverse.org/reference/group_by.html) to learn more about this.

3. If you only need sums or means, a somewhat faster alternative is using the base `R` functions `rowSums()` and `rowMeans()` in combination with `mutate()` (and possibly also `across()` plus selection helpers). For an explanation why this can be faster, you can read the [online documentation for `rowwise()`](https://dplyr.tidyverse.org/articles/rowwise.html).

---

## Aggregate variables

```{r rowwise-sum-output}
gp_covid %>% 
  select(hzcy044a:hzcy052a) %>% 
  glimpse()
```

---

## Aggregate variables

Rowwise transformations work the same way for means. Here, we create a mean score for the items that ask how much people trust specific people or institutions in dealing with the Corona virus. 

```{r rowwise-mean}
gp_covid <- 
  gp_covid %>%  
  rowwise() %>% 
  mutate(
    mean_trust = 
      mean(
        c_across(hzcy044a:hzcy052a), 
        na.rm = TRUE
      )
  ) %>% 
  ungroup()
```

---

## Aggregate variables

```{r rowwise-mean-output}
gp_covid %>% 
  select(hzcy044a:hzcy052a, mean_trust) %>% 
  glimpse()
```

---

## Conditional transformation

Sometimes, things are a bit more complicated. Simple recoding is insufficient when we need to base new variables based on the values of old variable(s).

---

## Simple conditional transformation

The simplest version of a conditional variable transformation is using an `ifelse()` statement.

```{r mutate-ifelse}
gp_covid <- 
  gp_covid %>% 
  mutate(
    high_education = 
      ifelse(education_cat == 3, "high", "not so high")
  )

gp_covid %>% 
  select(education_cat, high_education) %>% 
  sample_n(5)
```

.small[
*Note*: A more versatile option for creating dummy variables is the [`fastDummies` package](https://jacobkap.github.io/fastDummies/).
]

---

## Advanced conditional transformation

For more flexible (or complex) conditional transformations, the `case_when()` function from `dyplyr` is a powerful tool.

```{r case-when}
gp_covid <- 
  gp_covid %>% 
  mutate(
    pol_leaning_cat = 
      case_when(
        between(political_orientation, 0, 3) ~ "left",
        between(political_orientation, 4, 7) ~ "center",
        political_orientation > 7 ~ "right"
      )
  )

gp_covid %>% 
  select(political_orientation, pol_leaning_cat) %>% 
  sample_n(5)
```

---

## Conditional transformation based on multiple values

```{r case-when-2}
gp_covid <- 
  gp_covid %>% 
  mutate(
    pol_leaning_edu = 
      case_when(
        between(political_orientation, 0, 3) & high_education == "high" ~ "left high",
        between(political_orientation, 4, 7) & high_education == "high" ~ "center high",
        political_orientation > 7 & high_education == "high" ~ "right high",
        TRUE ~ "not so high"
      )
  )

gp_covid %>% 
  select(political_orientation, high_education, pol_leaning_edu) %>% 
  sample_n(5)
```

---

## `dplyr::case_when()`

A few things to note about `case_when()`:
- you can have multiple conditions per value
- conditions are evaluated consecutively
- when none of the specified conditions are met for an observation, by default, the new variable will have a missing value `NA` for that case
- if you want some other value in the new variables when the specified conditions are not met, you need to add `TRUE ~ value` as the last argument of the `case_when()` call
- to explore the full range of options for `case_when()` check out its [online documentation](https://dplyr.tidyverse.org/reference/case_when.html) or run `?case_when()` in `R`/*RStudio*

---

## `dplyr::case_when()`

```{r, case-when-cartoon, out.width = "95%", echo = F}
include_picture("dplyr_case_when.png")
```
<small><small>Artwork by [Allison Horst](https://github.com/allisonhorst/stats-illustrations)</small></small>

---

## Get a bit more programmy

So far, all of the previous tasks share two characteristics
- based on the structure of the whole dataset
- the output is again the whole dataset

Particularly in data analysis, our aim is often to extract information from a dataset (e.g., summary statistics, regression estimates).
