---
title: "Introduction to R for Data Analysis"
subtitle: "Appendix: Relational Data"
author: "Johannes Breuer & Stefan Jünger"
date: "2021-08-02"
presenter: Johannes
---
layout: true 

```{r child = "../config/sessions_setup.Rmd"}
```

---

## Relational data

In some cases, you may need to combine different data sets (e.g., if you have seperate data sets from waves of a longitudinal survey study or data on the same subjects from different sources). 

The simplest case is that the dataframes you want to combine either have the same cases/rows in the same order but different variables/columns, or the same columns/variables (with the same names) but different cases/rows.

In those cases, you can use the `base R` functions `rbind()` to add/combine rows or `cbind()` to add/combine columns, or, better yet, `bind_rows()` or `bind_cols()` from `dplyr`.

---

## Relational data

For all other situations, you can use the `merge()` function from `base R`. However, a more convenient option is using the the so-called *two-table verbs* from `dplyr` which allow you to combine two (or more) tabular data sets in various ways. We will look at two categories of those verbs in the following:

1. Mutating joins  

2. Filtering joins  

If you have ever worked with (or at least seen) `SQL` code, many of the following things will look familiar. For a more in-depth introduction, you can have a look at the [chapter on relational data](https://r4ds.had.co.nz/relational-data.html) in *R for Data Science*.

---

## Superhero data `r ji("bat")``r ji("zap")``r ji("spider_web")``r ji("merman")`

As we cannot use the data from the *GESIS Panel Special Survey on the Coronavirus SARS-CoV-2 Outbreak in Germany* to illustrate the different joins from `dlpyr`, we will, instead, use data from the [Super Heroes dataset from *Kaggle*](https://www.kaggle.com/claudiodavi/superhero-set).

```{r superhero-data, include = F}
heroes <- read_csv("./data/heroes_information.csv") %>% 
  select(name, Alignment, Race, Publisher) %>% 
  rename(Name = name) %>%
  mutate(Publisher = str_replace(Publisher, " Comics", "")) %>%
    filter(
    (Name == "Batgirl" & Race != "-") |
      Name == "Aquaman" |
      Name == "Catwoman" |
      Name == "Magneto" |
      Name == "Deadpool" |
      Name == "Poison Ivy" |
      Name == "Hellboy"
  )

publishers <- read_csv("./data/heroes_information.csv") %>% 
  select(Publisher) %>%
  distinct() %>% 
  filter(
    Publisher == "Marvel Comics" |
    Publisher == "DC Comics" |
    Publisher == "Image Comics"
  ) %>% 
  distinct() %>%
  mutate(Publisher = str_replace(Publisher, " Comics", ""),
    Founded = case_when(
    Publisher == "Marvel" ~ 1939L,
    Publisher == "DC" ~ 1934L,
    Publisher == "Image" ~ 1992L
  ),
  Location = case_when(
    Publisher == "Marvel" ~ "NYC (NY)",
    Publisher == "DC" ~ "Burbank (CA)",
    Publisher == "Image" ~ "Berkeley (CA)"
  ))

```

.pull-left[
.center[
**Superheroes**
```{r superhero-table, echo = F}
heroes %>%
  kable("html") %>% 
  kable_styling(font_size = 16)
```
]
]

.pull-right[
.center[
**Publishers**
```{r publishers-table, echo = F}
publishers %>%
  kable("html") %>% 
  kable_styling(font_size = 16)
```
]
]

---

## Mutating joins

.large[
> A mutating join [...] .highlight[first matches observations by their keys], then .highlight[copies across variables from one table to the other] ([R for Data Science](https://r4ds.had.co.nz/relational-data.html#mutating-joins)).

- `inner_join()`  

- `left_join()`  

- `right_join()`  

- `full_join()`
]

---

## Inner join

 > All rows from `x` where there are matching values in `y`, and all columns from `x` and `y`.

```{r inner-gif, echo=FALSE, out.width=".50%"}
woRkshoptools::include_picture("./inner-join.gif")
```
.small[
Animation by [Garrick Aden-Buie](https://github.com/gadenbuie/tidyexplain)
]

---

## Inner join example

.pull-left[
.center[
```{r superhero-table 2, echo = F}
heroes %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]

.pull-right[
.center[
```{r publishers-table-2, echo = F}
publishers %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]
<br>
```{r inber-join-display, eval = F}
heroes %>% 
  inner_join(publishers, by = "Publisher")
```

```{r inner-join-result, echo = F}
heroes %>% 
  inner_join(publishers, by = "Publisher") %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```

---

## Left join

> All rows from `x`, and all columns from `x` and `y`. Rows in `x` with no match in `y` will have `NA` values in the new columns.

```{r left-gif, echo=FALSE, out.width=".50%"}
woRkshoptools::include_picture("./left-join.gif")
```
.small[
Animation by [Garrick Aden-Buie](https://github.com/gadenbuie/tidyexplain)
]

---

## Left join example

.pull-left[
.center[
```{r superhero-table-3, echo = F}
heroes %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]

.pull-right[
.center[
```{r publishers-table-3, echo = F}
publishers %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]
<br>
```{r left-join-display, eval = F}
heroes %>% 
  left_join(publishers, by = "Publisher")
```

```{r left-join-result, echo = F}
heroes %>% 
  left_join(publishers, by = "Publisher") %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```

---

## Right join

> All rows from y, and all columns from `x` and `y`. Rows in `y` with no match in `x` will have `NA` values in the new columns.

```{r right-gif, echo=FALSE, out.width=".50%"}
woRkshoptools::include_picture("./right-join.gif")
```
.small[
Animation by [Garrick Aden-Buie](https://github.com/gadenbuie/tidyexplain)
]

---

## Right join example

.pull-left[
.center[
```{r superhero-table-4, echo = F}
heroes %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]

.pull-right[
.center[
```{r publishers-table-4, echo = F}
publishers %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]
<br>
```{r right-join-display, eval = F}
heroes %>% 
  right_join(publishers, by = "Publisher")
```

```{r right-join-result, echo = F}
heroes %>% 
  right_join(publishers, by = "Publisher") %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```

---

## Full join

> All rows and all columns from both `x` and `y`. Where there are not matching values, returns `NA` for the one missing.

```{r full-gif, echo=FALSE, out.width=".50%"}
woRkshoptools::include_picture("./full-join.gif")
```
.small[
Animation by [Garrick Aden-Buie](https://github.com/gadenbuie/tidyexplain)
]

---

## Full join example

.pull-left[
.center[
```{r superhero-table-5, echo = F}
heroes %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]

.pull-right[
.center[
```{r publishers-table-5, echo = F}
publishers %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]
<br>
```{r full-join-display, eval = F}
heroes %>% 
  full_join(publishers, by = "Publisher")
```

```{r full-join-result, echo = F}
heroes %>% 
  full_join(publishers, by = "Publisher") %>%
  kable("html") %>% 
  kable_styling(font_size = 9)
```

---

## Filtering joins

.large[
> Filtering joins .highlight[match observations in the same way as mutating joins], but .highlight[affect the observations, not the variables] ([R for Data Science](https://r4ds.had.co.nz/relational-data.html#filtering-joins)).

- `semi_join()`  

- `anti_join()`
]

---

## Semi join

> All rows from `x` where there are matching values in `y`, keeping just columns from `x`.

```{r semi-gif, echo=FALSE, out.width=".50%"}
woRkshoptools::include_picture("./semi-join.gif")
```
.small[
Animation by [Garrick Aden-Buie](https://github.com/gadenbuie/tidyexplain)
]

---

## Semi join example

.pull-left[
.center[
```{r superhero-table-6, echo = F}
heroes %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]

.pull-right[
.center[
```{r publishers-table-6, echo = F}
publishers %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]
<br>
```{r semi-join-display, eval = F}
heroes %>% 
  semi_join(publishers, by = "Publisher")
```

```{r semi-join-result, echo = F}
heroes %>% 
  semi_join(publishers, by = "Publisher") %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```

---

## Anti join

> All rows from `x` where there are not matching values in `y`, keeping just columns from `x`.

```{r anti-gif, echo=FALSE, out.width=".50%"}
woRkshoptools::include_picture("./anti-join.gif")
```
.small[
Animation by [Garrick Aden-Buie](https://github.com/gadenbuie/tidyexplain)
]

---

## Anti join example

.pull-left[
.center[
```{r superhero-table-7, echo = F}
heroes %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]

.pull-right[
.center[
```{r publishers-table-7, echo = F}
publishers %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
]
]
<br>
```{r anti-join-display, eval = F}
heroes %>% 
  anti_join(publishers, by = "Publisher")
```

```{r anti-join-result, echo = F}
heroes %>% 
  anti_join(publishers, by = "Publisher") %>%
  kable("html") %>% 
  kable_styling(font_size = 10)
```
