---
title: "Introduction to R for Data Analysis"
subtitle: "Data Visualization - Part 2"
author: "Johannes Breuer & Stefan Jünger"
date: "2021-08-05"
presenter: Stefan
editor_options: 
  chunk_output_type: console
---

layout: true 

```{r child = "../config/sessions_setup.Rmd"}
```

```{r load-gp-covid, include = FALSE}
gp_covid <- 
  haven::read_sav(
    "./data/ZA5667_v1-1-0.sav"
  ) %>% 
  sjlabelled::set_na(na = c(-1:-99, 97, 98)) %>% 
  rowwise() %>%
  mutate(
    mean_trust = 
      mean(
        c_across(hzcy044a:hzcy052a),
        na.rm = TRUE
      )
  ) %>% 
  ungroup() %>% 
  sjlabelled::remove_all_labels() %>% 
  mutate(
    pol_leaning_cat = 
      case_when(
        between(political_orientation, 0, 3) ~ "left",
        between(political_orientation, 4, 7) ~ "center",
        political_orientation > 7 ~ "right"
      ) 
  ) %>% 
  filter(pol_leaning_cat != "NA")
```

---

## Content of this session

.pull-left[
**What we will do**
- Visual checks of model assumptions
- Plotting coefficients
- Plotting predictions
- Plotting multiple models
]

.pull-right[
**What we won't do**
- Again, no crazy models
- No bayesian statistics
]

---

## Packages in this sessions

.pull-left[
We will again rely on packages from the `easystats` collection
- `paramters`
- `performance`
- **`see`**
]

.pull-right[
```{r easystats-pic, echo = FALSE}
woRkshoptools::include_picture("logo_wall.png")
```
]

.pull-left[
While it's straightforward to use, it still has some limitations. From time to time we will therefore also switch to the `sjPlot` package
]

.pull-right[
```{r sjPlot-pic, echo = FALSE}
woRkshoptools::include_picture("sjplot_logo.png")
```
]

---

## Estimating a linear model (OLS)

```{r linear-model}
linear_model <-
  lm(
    mean_trust ~
      age_cat + sex + education_cat + pol_leaning_cat,
    data = gp_covid
  )

library(parameters)

model_parameters(linear_model)
```

---

## Quick inspection using `base R`

.pull-left[
```{r lm_summary_plot, eval = FALSE}
par(mfrow = c(2, 2))
plot(linear_model)
```
]

.pull-right[
```{r ref.label = "lm_summary_plot", echo = FALSE}
```
]

---

## Using `easystats`

.pull-left[
```{r easystats-performance, eval = FALSE}
library(performance)
library(see)

check_normality(linear_model) %>% 
  plot()
```
]

.pull-right[
```{r ref.label = "easystats-performance", echo = FALSE}
```
]

---

## QQ-Plot (`easystats`)

.pull-left[
```{r easystats-qq, eval = FALSE}
check_normality(linear_model) %>% 
  plot(type = "qq")
```
]

.pull-right[
```{r ref.label = "easystats-qq", echo = FALSE}
```
]

---

## Heteroscedasticity (`easystats`)

.pull-left[
```{r easystats-hetero, eval = FALSE}
check_heteroscedasticity(linear_model) %>% 
  plot()
```
]

.pull-right[
```{r ref.label = "easystats-hetero", echo = FALSE}
```
]

---

## Check outliers (`easystats`)

.pull-left[
```{r check-outliers, eval = FALSE}
check_outliers(linear_model) %>% 
  plot()
```
]

.pull-right[
```{r ref.label = "check-outliers", echo = FALSE}
```
]


---

## Coefficient plot (`easystats`)

.pull-left[
```{r easystats-coef-plot, eval = FALSE}
library(parameters)

model_parameters(linear_model) %>% 
  plot()
```
]

.pull-right[
```{r ref.label = "easystats-coef-plot", echo = FALSE}
```
]

---

## Changing labels: it's `ggplot2`-based

.pull-left[
```{r labels-change, eval = FALSE}
library(ggplot2)

model_parameters(linear_model) %>% 
  plot() +
  scale_x_discrete(
    labels = # in reversed order!
      c(
        "Pol. Right (Ref. Center)",
        "Pol. Left (Ref. Center)", 
        "Education", 
        "Sex", 
        "Age"
      )
  )
```
]

.pull-right[
```{r ref.label = "labels-change", echo = FALSE}
```
]

---

## Making it more dim and scientific

.pull-left[
```{r labels-change-2, eval = FALSE}
library(ggplot2)

model_parameters(linear_model) %>% 
  
  plot() +
  scale_x_discrete(
    labels = # in reversed order!
      c(
        "Pol. Right (Ref. Center)",
        "Pol. Left (Ref. Center)", 
        "Education", 
        "Sex", 
        "Age"
      )
  ) +
  scale_colour_grey(start = 0, end = 0) + 
    guides(color = "none")
```
]

.pull-right[
```{r ref.label = "labels-change-2", echo = FALSE}
```
]

---

## Predictions (`sjPlot`)

.pull-left[
```{r predictions, eval = FALSE}
library(sjPlot)

linear_model %>% 
  plot_model(
    type = "pred", 
    terms = "age_cat"
  )

```
]

.pull-right[
```{r ref.label = "predictions", echo = FALSE}
```
]

---

## Again, it's `ggplot2`-based

.pull-left[
```{r predictions-nicer, eval = FALSE}
linear_model %>% 
  plot_model(
    type = "pred", 
    terms = "age_cat"
  ) +
  xlab("Age") +
  ylab("Trust") +
  ggtitle("Linear Model") +
  theme_bw()
```
]

.pull-right[
```{r ref.label = "predictions-nicer", echo = FALSE}
```
]

---

## Estimating multiple models

```{r median-split}
gp_covid <-
  gp_covid %>% 
  mutate(
    mean_trust_median = 
      ifelse(
        mean_trust <= median(mean_trust, na.rm = TRUE),
        0,
        1
      )
  )
```

---

## Logistic regressions

```{r log-reg}
logistic_model <-
  glm(
    mean_trust_median ~
      age_cat + sex + education_cat + pol_leaning_cat,
    data = gp_covid,
    family = binomial(link = "logit"),
  )

model_parameters(logistic_model)
```

---

## Linear probability model

```{r prob-reg}
linear_probability_model <-
  lm(
    mean_trust_median ~
      age_cat + sex + education_cat + pol_leaning_cat,
    data = gp_covid,
  )

model_parameters(linear_probability_model)
```

---

## Comparing model fits (`easystats`)

.pull-left[
```{r comparing-model-fit, eval = FALSE}
library(performance)

compare_performance(
  linear_model,
  logistic_model,
  linear_probability_model
) %>% 
  plot()
```
]

.pull-right[
```{r ref.label = "comparing-model-fit", echo = FALSE}
```
]

---

## Comparing model parameters (`sjPlot`)

.pull-left[
```{r comparing-parameters, eval = FALSE}
plot_models(
  linear_model,
  logistic_model,
  linear_probability_model,
  std.est = "std"
)
```
]

.pull-right[
```{r ref.label = "comparing-parameters", echo = FALSE}
```
]

---

## Adjusting it within the function

.pull-left[
```{r comparing-parameters-adjusted, eval = FALSE}
plot_models(
  linear_model,
  logistic_model,
  linear_probability_model,
  std.est = "std",
  m.labels = 
    c(
      "Linear",
      "Logistic", 
      "Linear Probability"
      ),
  axis.labels = 
      c(
        "Pol. Right (Ref. Center)",
        "Pol. Left (Ref. Center)", 
        "Education", 
        "Sex", 
        "Age"
      )
) 
```
]

.pull-right[
```{r ref.label = "comparing-parameters-adjusted", echo = FALSE}
```
]

---

## But it's still a `ggplot2` object

.pull-left[
```{r comparing-parameters-adjusted-bw, eval = FALSE}
plot_models(
  linear_model,
  logistic_model,
  linear_probability_model,
  # std.est = "std",
  m.labels = 
    c(
      "Linear",
      "Logistic", 
      "Linear Probability"
      ),
  axis.labels = 
      c(
        "Pol. Right (Ref. Center)",
        "Pol. Left (Ref. Center)", 
        "Education", 
        "Sex", 
        "Age"
      )
) +
  theme_bw()
```
]

.pull-right[
```{r ref.label = "comparing-parameters-adjusted-bw", echo = FALSE}
```
]

---

## Plotting predictions from multiple models

```{r extract-linear-predictions}
linear_predictions <-
  get_model_data(
    linear_model,
    term = "age_cat",
    type = "pred"
  )

linear_predictions
```

---

## Predictions from the other two models

.pull-left[
```{r extract-logistic-predictions}
logistic_predictions <-
  get_model_data(
    logistic_model,
    term = "age_cat",
    type = "pred"
  )
```
]

.pull-right[
```{r extract-lp-predictions}
linear_probability_predictions <-
  get_model_data(
    linear_probability_model,
    term = "age_cat",
    type = "pred"
  )
```
]

---

## Combining it in one table

```{r table}
library(dplyr)
library(tibble)

all_predictions <-
  bind_rows(
    linear_predictions %>% 
      mutate(model = "Linear Model"),
    logistic_predictions %>% 
      mutate(model = "Logistic Model"),
    linear_probability_predictions %>% 
      mutate(model = "Linear Probability Model")
  ) %>% 
  as_tibble()
```

---

## Plotting them with `facet_wrap()`

.pull-left[
```{r all-in-one, eval = FALSE}
ggplot(
  all_predictions, 
  aes(x = x, y = predicted)
  ) +
  geom_line() +
  geom_line(aes(y = conf.high), linetype = "dashed") +
  geom_line(aes(y = conf.low), linetype = "dashed") +
  facet_wrap(~model) +
  theme_bw()
```
]

.pull-right[
```{r ref.label = "all-in-one", echo = FALSE}
```
]

---

## Plotting only the binary models

.pull-left[
```{r only-two, eval = FALSE}
only_two <-
  ggplot(
    all_predictions %>% 
      filter(model != "Linear Model"), 
    aes(x = x, y = predicted)
  ) +
  geom_line() +
  geom_line(aes(y = conf.high), linetype = "dashed") +
  geom_line(aes(y = conf.low), linetype = "dashed") +
  facet_wrap(~model) +
  theme_bw()

only_two
```
]

.pull-right[
```{r ref.label = "only-two", echo = FALSE}
```
]

---

## The same logic for real AME

```{r margins}
library(margins)

logistic_model_AME <-
  cplot(
    logistic_model,
    what = "prediction",
    draw = FALSE
  )

linear_probability_model_AME <-
  cplot(
    linear_probability_model,
    what = "prediction",
    draw = FALSE
  )
```

---

## Combining AME data

```{r combine-data-again}
all_AME <-
  bind_rows(
    logistic_model_AME %>% 
      mutate(model = "Logistic Model"),
    linear_probability_model_AME %>% 
      mutate(model = "Linear Probability Model")
  ) %>% 
  as_tibble()
```

---

## Plotting them again

.pull-left[
```{r only-two-AME, eval = FALSE}
only_two_AME <- 
  ggplot(
    all_AME, 
    aes(x = xvals, y = yvals)
  ) +
  geom_line() +
  geom_line(aes(y = upper), linetype = "dashed") +
  geom_line(aes(y = lower), linetype = "dashed") +
  facet_wrap(~model) +
  theme_bw()

only_two_AME
```
]

.pull-right[
```{r ref.label = "only-two-AME", echo = FALSE}
```
]

---

## Combining them with previous predictions

.pull-left[
```{r patchwork-combining, eval = FALSE}
library(patchwork)

(only_two +
  ggtitle("Simple Predictions") +
    ylab("Predicted Trust Value") +
    xlab("Age")) / 
  (only_two_AME +
  ggtitle("AME Predictions") +
    ylab("Predicted Trust Value") +
    xlab("Age"))
  

```
]

.pull-right[
```{r ref.label = "patchwork-combining", echo = FALSE}
```
]




