---
title: "Introduction to R for Data Analysis"
subtitle: "Outlook"
author: "Johannes Breuer & Stefan Jünger"
date: "2021-08-06"
presenter: Stefan
editor_options: 
  chunk_output_type: console
---

layout: true 

```{r child = "../config/sessions_setup.Rmd"}
```

---

## Course schedule - Day 1

```{r schedule-1, echo = F}
course_content_1
```

---

## Course schedule - Day 2

```{r schedule-2, echo = F}
course_content_2
```

---

## Course schedule - Day 3

```{r schedule-3, echo = F}
course_content_3
```

---

## Course schedule - Day 4

```{r schedule-4, echo = F}
course_content_4
```

---

## Course schedule - Day 5

```{r schedule-5, echo = F}
course_content_5
```

---

## Where to go from here?

Hopefully, after this week, you feel prepared to do your next steps in `R` alone.

Our recommendations for the future:

- Keep up working with `R`!
- If time permits, do stuff you usually do in `SPSS` or `Stata` in `R`, even when it's harder
- Try to do at least one research task solely in `R` (one analysis, a whole paper, a report, etc.)
- Search for solutions on the internet
  - there's way more than for `SPSS` or `Stata`

---

## Working with other data types

Both Johannes and Stefan use different data types in our daily work.
- digital behavioral data (Johannes)
- georeferenced/geospatial data (Stefan)

**`R` is data agnostic! Thus, `R` can serve you as a fancy data science tool to extract tweets from Twitter but also as a full-blown Geographic Information System (GIS)**


.footnote[* https://github.com/jobreu/twitter-linking-workshop-2021]
.footnote[** https://github.com/StefanJuenger/gesis-workshop-geospatial-techniques-R]

---

## Geospatial data

.pull-left[
```{r mapsf, eval = FALSE}
library(mapsf)

mtq <- mf_get_mtq()

mf_map(x = mtq)
mf_map(x = mtq, var = "POP", type = "prop")
mf_layout(
  title = "Population in Martinique", 
  credits = "T. Giraud; Sources: INSEE & IGN, 2018"
)
```
]

.pull-right[
```{r mapsf-print, echo = FALSE}
library(mapsf)

mtq <- mf_get_mtq()

mf_map(x = mtq)
mf_map(x = mtq, var = "POP", type = "prop")
mf_layout(
  title = "Population in Martinique", 
  credits = "T. Giraud; Sources: INSEE & IGN, 2018"
)
```
]

Example from: https://riatelab.github.io/mapsf/


---

## 'Web development' using `R`

These days, a lot of `R` packages provides tools originally developed for the web. For example:
- [bookdown](https://cran.r-project.org/web/packages/bookdown/index.html) enables you to publish your book written in `R Markdown` online
- [pkgdown](https://cran.r-project.org/web/packages/pkgdown/index.html) does the same for your own `R` package
- [blogdown](https://cran.r-project.org/web/packages/blogdown/index.html) is more general and helps you with creating websites (see below)

---

## Shiny apps

> Shiny is an R package that makes it easy to build interactive web apps straight from R. You can host standalone apps on a webpage or embed them in R Markdown documents or build dashboards. You can also extend your Shiny apps with CSS themes, htmlwidgets, and JavaScript actions.

https://shiny.rstudio.com/

---

## Example 1:

---

## Example 2:

---

## Creating your own homepage

.pull-left[
```{r home-johannes, echo = FALSE}
woRkshoptools::include_picture("homepage_johannes.png")
```
.center[.small[https://www.johannesbreuer.com/]]
]

.pull-right[
```{r home-stefan, echo = FALSE}
woRkshoptools::include_picture("homepage_stefan.png")
```
.center[.small[https://stefanjuenger.github.io/]]
]

---

## Acknowledgements `r ji("heart")`

All slides were created with the `R` package [`xaringan`](https://github.com/yihui/xaringan) which builds on [`remark.js`](https://remarkjs.com), [`knitr`](http://yihui.name/knitr), and [`RMarkdown`](https://rmarkdown.rstudio.com). The exercises were created with the [`unilur` package](https://github.com/koncina/unilur).  

Please make sure to properly cite all data that you use for your research (archives usually provide suggested citations). Also make sure to cite the free and open-source software (FOSS) that you use, such as `R` and the packages for it. To know how to do that, you can use the function `citation(packagename)` in `R`.

We thank the GESIS Summer School 2021 organizers for inviting us and all of you for participating!