---
title: "Introduction to R for Data Analysis"
subtitle: "Outlook"
author: "Johannes Breuer & Stefan Jünger"
date: "2021-08-06"
presenter: Stefan
editor_options: 
  chunk_output_type: console
---

layout: true 

```{r child = "../config/sessions_setup.Rmd"}
```

---

## Course schedule - Day 1

```{r schedule-1, echo = F}
course_content_1
```

---

## Course schedule - Day 2

```{r schedule-2, echo = F}
course_content_2
```

---

## Course schedule - Day 3

```{r schedule-3, echo = F}
course_content_3
```

---

## Course schedule - Day 4

```{r schedule-4, echo = F}
course_content_4
```

---

## Course schedule - Day 5

```{r schedule-5, echo = F}
course_content_5
```

---

## Where to go from here?

---

## Working with other data types

Both Johannes and Stefan use different data types in our daily work.
- digital behavioral data (Johannes)
- georeferenced/geospatial data (Stefan)

**`R` is data agnostic! Thus, `R` can serve you as a fancy data science tool to extract tweets from Twitter^ but also as a full-blown Geographic Information System (GIS)^2**

.small[
^1 https://github.com/jobreu/twitter-linking-workshop-2021
^2 https://github.com/StefanJuenger/gesis-workshop-geospatial-techniques-R
]

---

## 'Web development' using `R`

---

## Shiny apps

---

## Creating your own homepage

.pull-left[
```{r home-johannes, echo = FALSE}
woRkshoptools::include_picture("homepage_johannes.png")
```
.center[.small[https://www.johannesbreuer.com/]]
]

.pull-right[
```{r home-stefan, echo = FALSE}
woRkshoptools::include_picture("homepage_stefan.png")
```
.center[.small[https://stefanjuenger.github.io/]]
]

---

## Acknowledgements `r ji("heart")`

All slides were created with the `R` package [`xaringan`](https://github.com/yihui/xaringan) which builds on [`remark.js`](https://remarkjs.com), [`knitr`](http://yihui.name/knitr), and [`RMarkdown`](https://rmarkdown.rstudio.com). The exercises were created with the [`unilur` package](https://github.com/koncina/unilur).  

Please make sure to properly cite all data that you use for your research (archives usually provide suggested citations). Also make sure to cite the free and open-source software (FOSS) that you use, such as `R` and the packages for it. To know how to do that, you can use the function `citation(packagename)` in `R`.

We thank the ESRA 2021 organizers (especially Marieke Haan) for inviting us, Hidde van Beek for the support during the workshop, and all of you for participating!