---
title: "Introduction to R for Data Analysis"
subtitle: "Outlook"
author: "Johannes Breuer & Stefan Jünger"
date: "2021-08-06"
presenter: Stefan
editor_options: 
  chunk_output_type: console
---

layout: true 

```{r child = "../config/sessions_setup.Rmd"}
```

---

## Recap: Course schedule - Day 1

```{r schedule-1, echo = F}
course_content_1
```

---

## Recap: Course schedule - Day 2

```{r schedule-2, echo = F}
course_content_2
```

---

## Recap: Course schedule - Day 3

```{r schedule-3, echo = F}
course_content_3
```

---

## Recap: Course schedule - Day 4

```{r schedule-4, echo = F}
course_content_4
```

---

## Recap: Course schedule - Day 5

```{r schedule-5, echo = F}
course_content_5
```

---

## Where to go from here?

Hopefully, after this week, you feel prepared to do your next steps in `R`.

Some recommendations for continuing your jou`R`ney:

- Keep up working with `R`!
- If time permits, do stuff you usually do in `SPSS` or `Stata` in `R`, even when it's harder
- Try to do at least one research task solely in `R` (one analysis, a whole paper, a report, etc.)
- Look for tutorials and guides online
  - trust us, there's way more (good & free) online material for `R` than there is, e.g., for `SPSS` or `Stata`
  
---

## You voted for: Multilevel models

Multilevel/Hierarchical or Mixed Regression Models (let's stick with MLM) have been quite popular for some time. They allow the incorporation of
- random intercepts
- random slopes

Simply put: regression coefficients that vary across groups.

---

## MLM in `R`

Of course, you can estimate such models in `R` as well. There are packages for that:
- [`lme4`](https://cran.r-project.org/web/packages/lme4/index.html) may the most popular one
  - provides functions to estimate linear models (`lmer()`) or generalized linear models (`glmer()`)
- ...
- but there are others, e.g., for ordinal responses like the `clmm2()` function from the [`ordinal`](https://cran.r-project.org/web/packages/ordinal/index.html) package

---

## Toy example

Let's start with loading the data and creating fictional clusters.

```{r read-gp}
# load data
gp_covid <- readRDS("./data/corona_survey.rds")

# simulate clusters
gp_covid <-
  gp_covid %>% 
  mutate(
    clusters = sample(1:50, n(), replace = TRUE)
  )

table(gp_covid$clusters)
```

---

## Re-running models from the course as MLM: random intercept

```{r run-mlm}
library(lme4)

mlm <-
  lmer(
    risk_self ~ 1 + left_right + (1 | clusters),
    data = gp_covid
    )
```

---

## Standard Output

.small[
```{r mlm-summary}
summary(mlm)
```
]

---

## Output using the `parameters` package

.small[
```{r mlm-parameters}
library(parameters)

model_parameters(mlm)
```
]

---

## Adding a random slope

```{r run-mlm-rs}
library(lme4)

mlm_2 <-
  lmer(
    risk_self ~ 1 + left_right + (1 + left_right | clusters),
    data = gp_covid
    )
```

---

## Output using the `parameters` package

.small[
```{r mlm-parameters-2}
model_parameters(mlm_2)
```
]

---

## btw: You can also plot MLM with `easystats`

.pull-left[
```{r plot-mlm, eval = FALSE}
model_parameters(mlm) %>% 
  plot()
```
]

.pull-right[
```{r ref.label = "plot-mlm", echo = FALSE, out.width = "95%"}
```
]

---

## ...or `sjPlot` for predictions

.pull-left[
```{r plot-mlm-sjplot, eval = FALSE}
library(sjPlot)
plot_model(mlm, type = "pred")
```
]

.pull-right[
```{r ref.label = "plot-mlm-sjplot", echo = FALSE, out.width = "95%"}
```
]

---

## Working with other data types

Both Johannes and Stefan use different data types in their daily work:
- digital trace data (Johannes) <sup>*</sup>
- georeferenced/geospatial data (Stefan)<sup>**</sup>

**Remember that `R` is data-agnostic! It can serve as a fancy data science tool for extracting social media data but also as a full-blown Geographic Information System (GIS)**


.footnote[
<sup>*</sup> see, e.g., https://github.com/jobreu/twitter-linking-workshop-2021

<sup>**</sup> see, e.g., https://github.com/StefanJuenger/gesis-workshop-geospatial-techniques-R
]

---

## What Are Geospatial Data?

.pull-left[
Data with a direct spatial reference $\rightarrow$ **geo-coordinates**
- Information about geometries
- Optional: Content in relation to the geometries

Can be projected jointly in one single space
- Allows data linking and extraction of substantial information
]

.pull-right[
```{r pic-geometries-1, echo = FALSE, out.width = "85%"}
woRkshoptools::include_picture("fig_geometries.png")
```

.tinyisher[Sources: OpenStreetMap / GEOFABRIK (2018), City of Cologne (2014), and the Statistical Offices of the Federation and the Länder (2016) / Jünger, 2019]
]

---

## Mapping is so easy nowadays

.pull-left[
```{r mapsf, eval = FALSE}
library(mapsf)

mtq <- mf_get_mtq()

mf_map(x = mtq)
mf_map(x = mtq, var = "POP", type = "prop")
mf_layout(
  title = "Population in Martinique", 
  credits = "T. Giraud; Sources: INSEE & IGN, 2018"
)
```
]

.pull-right[
```{r mapsf-print, echo = FALSE}
library(mapsf)

mtq <- mf_get_mtq()

mf_map(x = mtq)
mf_map(x = mtq, var = "POP", type = "prop")
mf_layout(
  title = "Population in Martinique", 
  credits = "T. Giraud; Sources: INSEE & IGN, 2018"
)
```
]

Example from: https://riatelab.github.io/mapsf/

---

## Interactive Mapping!

.pull-left[
```{r mapview, eval = FALSE}
library(mapview)

mapview(mtq["POP"])
```
]

.pull-right[
```{r mapview-print, echo = FALSE, eval = FALSE}
library(mapview)

mapview(mtq["POP"])
```
]

---

## 'Web development' using `R`

These days, a lot of `R` packages provides tool originally developed for the web. For example:
- [bookdown](https://cran.r-project.org/web/packages/bookdown/index.html) enables you to publish your book written in `R Markdown` online
- [pkgdown](https://cran.r-project.org/web/packages/pkgdown/index.html) does the same for your own `R` package
- [blogdown](https://cran.r-project.org/web/packages/blogdown/index.html) is more general and helps you with creating websites (example to follow)

---

## Shiny apps

> Shiny is an R package that makes it easy to build interactive web apps straight from R. You can host standalone apps on a webpage or embed them in R Markdown documents or build dashboards. You can also extend your Shiny apps with CSS themes, htmlwidgets, and JavaScript actions.

https://shiny.rstudio.com/

---

class: middle

## Example 1: Movie Explorer

.center[https://shiny.rstudio.com/gallery/movie-explorer.html]

---

class: middle

## Example 2: CRAN explorer

.center[https://gallery.shinyapps.io/cran-explorer/]

---

## Creating your own homepage with `R`

.pull-left[
```{r home-johannes, echo = FALSE}
woRkshoptools::include_picture("homepage_johannes.png")
```
.center[.small[https://www.johannesbreuer.com/]]
]

.pull-right[
```{r home-stefan, echo = FALSE}
woRkshoptools::include_picture("homepage_stefan.png")
```
.center[.small[https://stefanjuenger.github.io/]]
]

.center[Powered by  [`blogdown`](https://cran.r-project.org/web/packages/blogdown/index.html) &[ Hugo Academic](https://academic-demo.netlify.app/)]

---

## Writing your own `R` packages

.pull-left[
At a certain point (not now!), you may want to consider writing your own `R` package
- useful for creating reproducible code
- great for distributing your work to others
- for example, we created an [`R` package](https://stefanjuenger.github.io/woRkshoptools/) to facilitate working on our workshop materials
]

.pull-right[
```{r r-packages, echo = FALSE}
woRkshoptools::include_picture("r_packages.jpg")
```
[Read the book here!](https://r-pkgs.org/)
]

---

class: middle

## It's straightforward in `RStudio`

```{r new-package, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("new_package.png")
```

---

## Acknowledgements `r ji("heart")`

All slides were created with the `R` package [`xaringan`](https://github.com/yihui/xaringan) which builds on [`remark.js`](https://remarkjs.com), [`knitr`](http://yihui.name/knitr), and [`RMarkdown`](https://rmarkdown.rstudio.com). The exercises were created with the [`unilur` package](https://github.com/koncina/unilur).  

Please make sure to properly cite all data that you use for your research (archives usually provide suggested citations). Also make sure to cite the free and open-source software (FOSS) that you use, such as `R` and the packages for it. To know how to do that, you can use the function `citation(packagename)` in `R`.

We thank the *GESIS Training* team for taking good care of the organization of this course (and the whole Summer School) and all of you for participating!