---
title: 'Exercise 1 Getting Started - Solutions'
# title: 'Exercise 1 Getting Started'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

As this is the very first exercise in this workshop it is quite easy and short. Its purpose is also to get used to this exercise format and, more importantly, to get a first feeling for working with `R` and *RStudio*.

*Note*: You can find the solutions for this exercise as well as all other exercises in the `solutions` folder in the repo/directory that contains the course materials. You can copy code from these exercise files by clicking on the small blue clipboard icon in the upper right corner of the code boxes.

The first thing we would like you to do in this exercise is to explore the *RStudio* interface a bit. Just "have a look around" and see, if you can find the elements and options we have discussed so far.

After you have made yourself a bit familiar with the *RStudio* interface, start using the console interactively by entering and executing the following simple calculations:

21 * 2

2^3

15/3

Feel free to play around a bit and try a few things out.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
The first simple task in this exercise is to assign the number 42 to an object named `the_answer` and then print the object to the console.
```

```{block, opts.label = "clues"}
You need to use the assignment operator `<-` for this.
```

```{r first.exercise, solution = TRUE}
the_answer <- 42

the_answer
```

Also have a look at the `Environment` tab in the *RStudio* interface. The object you created should now be visible there.

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Use a simple function to calculate the square root of 49.
```

```{r second-exercise, solution = TRUE}
sqrt(49)
```

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Check out the help file for the `grep()` function. How many required arguments does it have?
```

```{r third-exercise, solution = TRUE, eval = FALSE}
?grep

# It has two required arguments.
```
