---
title: 'Exercise 1 Confirmatory Analysis - Solutions'
# title: 'Exercise 1 Confirmatory Analysis'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

We are now entering the phase of doing serious statistics: we run regression models, inspect them, plot, and tabulate them.

For this exercise, let's pretend to be researchers interested in the determinants of the subjective risk of being treated for COVID-19 in a hospital. We suspect that age plays an important role, something we have already assessed by visualizing a similar relationship. So... we need the *GESIS Panel* COVID-19 survey data again:

```{r loadgpcovid, message=FALSE,}
library(dplyr)
library(haven)
library(sjlabelled)

gp_covid <-
  read_sav(
    "../data/ZA5667_v1-1-0.sav"
  ) %>% 
  set_na(na = c(-1:-99, 97)) %>% 
  mutate(
    likelihood_hospital = hzcy003a,
    age_cat = as.factor(age_cat)
  ) %>% 
  remove_all_labels()
```

We start our analysis by checking the relationship with an ANOVA. We also include some very essential covariates: gender, political orientation, and marital status. 

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Run an ANOVA on the proposed relationship between the subjective likelihood of treatment in a hospital and age. Also, include all the covariates.
```

```{block, opts.label = "clues"}
We had a similar analysis before... you can simply add the control variables with `+`.
```

```{r anova, solution = TRUE}
serious_anova <- 
  aov(
    likelihood_hospital ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid
  )

summary(serious_anova)
```

Alright, there's definitely something in the data we should investigate. Oftentimes this is considered bad practice, but we could try to create a median split for the dependent variable and run a logistic regression.

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Create a median split for the variable `likelihood_hospital` and run a logistic regression with all variables. What do you see in the results?
```

```{block, opts.label = "clues"}
You can create new variables fairly easily with the function `dplyr::mutate()` in combination with the `ifelse()` function. 
```

```{r mediancutlog, solution = TRUE}
gp_covid <-
  gp_covid %>% 
  mutate(
    likelihood_hospital_cut =
      ifelse(
        likelihood_hospital > median(likelihood_hospital, na.rm = TRUE),
        1,
        0
      )
  )

serious_logistic_regression <- 
  glm(
    likelihood_hospital_cut ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid,
    family = binomial(link = "logit")
  )

summary(serious_logistic_regression)

# Interpretation:
# With increasing age, the probability of thinking it's likely to be 
# hospitalized because of COVID-19 increases. But really?
```

Running one model is not enough. It may be that the assumptions for logistic regression are not fulfilled, or a reviewer simply doesn't like these types of regressions. Instead, she proposes to run a binomial regression but with a cauchit link.

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Run the same model with the sole difference of using a cauchit link. How would you interpret the different regression coefficients?
```

```{block, opts.label = "clues"}
Have a look at the help page `?family` to see how you can include a cauchit link.
```

```{r cauchit, solution = TRUE}
serious_cauchit_regression <- 
  glm(
    likelihood_hospital_cut ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid,
    family = binomial(link = "cauchit")
  )

summary(serious_cauchit_regression)

# Interpretation of difference:
# I don't know...
```

Using different link functions is sometimes done as they provide different model fits. That's definitely something we should investigate as well.

```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Compare both regression models with an ANOVA. Use the option `test = "LRT"` to perform a likelihood ratio test. What's your interpretation?
```

```{block, opts.label = "clues"}
A p-value considered as statistically significant would indicate a difference between the models.
```

```{r anovalrt, solution = TRUE}
anova(
  serious_logistic_regression, 
  serious_cauchit_regression,
  test = "LRT"
  )

# Interpretation of difference:
# It seems there is no big difference. We can make the reviewer happy with the
# cauchit models and still keep our results.
```


